The Diagnostic Potential & Interactive Dynamics of the Colorectal Cancer Virome

### 2018-09-30 doc/manuscript.Rmd

Human viruses (those that infect human cells) have been associated with many cancers, largely due to their mutagenic and functionally manipulative abilities. Despite this, cancer microbiome studies have almost exclusively focused on bacteria instead of viruses. We began evaluating the cancer virome by focusing on colorectal cancer, a primary cause of morbidity and mortality throughout the world, and a cancer linked to altered colonic bacterial community compositions but with an unknown association with the gut virome. We used 16S rRNA gene, whole shotgun metagenomic, and purified virus metagenomic sequencing of stool to evaluate the differences in human colorectal cancer virus and bacterial community composition. Through random forest modeling we identified differences in the healthy and colorectal cancer virome. The cancer-associated virome consisted primarily of temperate bacteriophages that were also predicted to be bacteria-virus community network hubs. These results provide foundational evidence that bacteriophage communities are associated with colorectal cancer and potentially impact cancer progression by altering the bacterial host communities.

@@

Colorectal cancer is a leading cause of cancer-related death in the United States and worldwide. Its risk and severity have been linked to colonic bacterial community composition. Although human-specific viruses have been linked to other cancers and diseases, little is known about colorectal cancer virus communities. We addressed this knowledge gap by identifying differences in colonic virus communities in the stool of colorectal cancer patients and how they compared to bacterial community differences. The results suggested an indirect role for the virome in impacting colorectal cancer by modulating their associated bacterial community. These findings both support a biological role for viruses in colorectal cancer and provide a new understanding of basic colorectal cancer etiology.

@@

# Introduction
The human gut virome is the community of all viruses found in the gut, including bacteriophages (viruses that only infect bacteria), eukaryotic viruses (viruses that only infect eukaryotic cells), and human-specific viruses (viruses that only infect human cells). Due to their mutagenic abilities and propensity for functional manipulation, human viruses are strongly associated with, and in many cases cause, cancer [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up]. Because bacteriophages are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk], and members of those bacterial communities have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb; @Arthur:2012kl], bacteriophages have the potential to indirectly impact cancer as well. The gut virome therefore has a potential to be associated with, and potentially impact, human cancer. Altered human virome composition and diversity have already been identified in diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], cystic fibrosis [@Willner:2009dq], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb]. The strong association of bacterial communities with colorectal cancer, the previous identification of human-specific viruses that cause cancer, and the precedent for the virome to impact other human diseases suggest that colorectal cancer may be associated with altered virus communities.

Colorectal cancer is the second leading cause of cancer-related deaths in the United States [@Siegel:2014jo]. The US National Cancer Institute estimates over 1.5 million Americans were diagnosed with colorectal cancer in 2016 and over 500,000 Americans died from the disease [@Siegel:2014jo]. Growing evidence suggests that an important component of colorectal cancer etiology may be perturbations in the colonic bacterial community [@Zackular:2016en; @Zackular:2014fba; @Baxter:2014hb; @Dejea:2014fz; @Arthur:2012kl]. Work in this area has led to a proposed disease model in which bacteria colonize the colon, develop biofilms, promote inflammation, and enter an oncogenic synergy with the cancerous human cells [@Flynn:2016iu]. This association also has allowed researchers to leverage bacterial community signatures as biomarkers to provide accurate, noninvasive colorectal cancer detection from stool [@Zackular:2014fba; @Baxter:2016dja; @Zeller:2014ix]. While an understanding of colorectal cancer bacterial communities has proven fruitful both for disease classification and for identifying the underlying disease etiology, bacteria are only a subset of the colon microbiome. Viruses are another important component of the colon microbial community that have yet to be studied in the context of colorectal cancer. We evaluated disruptions in virus and bacterial community composition in a human cohort whose stool was sampled at the three relevant stages of cancer development: healthy, adenomatous, and cancerous.

Colorectal cancer progresses in a stepwise process that begins when healthy tissue develops into a precancerous polyp (i.e., adenoma) in the large intestine [@Fearon:2011go]. If not removed, the adenoma may develop into a cancerous lesion that can invade and metastasize, leading to severe illness and death. Progression to cancer can be prevented when precancerous adenomas are detected and removed during routine screening [@Levin:2008kz; @Zauber:2015cr]. Survival for colorectal cancer patients may exceed 90% when the lesions are detected early and removed [@Levin:2008kz]. Thus, work that aims to facilitate early detection and prevention of progression beyond early cancer stages has great potential to inform therapeutic development.

Here we begin to address the knowledge gap of whether virus community composition is altered in colorectal cancer and, if it is, how those differences might impact cancer progression and severity. We also aimed to evaluate the virome's potential for use as a diagnostic biomarker. The implications of this study are threefold. *First*, this work supports a biological role for the virome in colorectal cancer development and suggests that more than the bacterial members of the associated microbial communities are involved in the process. *Second*, we present a supplementary virus-based approach for classification modeling of colorectal cancer using stool samples. *Third*, we provide initial support for the importance of studying the virome as a component of the microbiome ecological network, especially in cancer.

@@

## Unaltered Diversity in Colorectal Cancer
Microbiome and disease associations are often described as being of an altered diversity (i.e. "dysbiotic"). Therefore, we first evaluated the influence of colorectal cancer on virome OVU diversity. We evaluated differences in communities between disease states using the Shannon diversity, richness, and Bray-Curtis metrics. We observed no significant alterations in either Shannon diversity or richness in the diseased states as compared to the healthy state **(Figure \ref{betaogu} C-D)**. There was no statistically significant clustering of the disease groups (ANOSIM p-value `r pvalcalc(alphadf[c(alphadf$Title %in% "anovawithoutneg"),"Stat"])`, **Figure \ref{betaogu}**). Notably, there was a significant difference between the few blank controls that remained after rarefying the data and the other study groups (ANOSIM p-value `r pvalcalc(alphadf[c(alphadf$Title %in% "anovawithneg"),"Stat"])`, **Figure \ref{betaogunegative})**, further supporting the quality of the sample set. In summary, standard alpha and beta diversity metrics were insufficient for capturing virus community differences between disease states **(Figure \ref{betaogu})**. This is consistent with what has been observed when the same metrics were applied to 16S rRNA gene sequences and metagenomic samples [@Zeller:2014ix; @Zackular:2014fba; @Baxter:2016dja] and points to the need for alternate approaches to detect the impact of colorectal cancer disease state on these community structures.

## Virome Composition in Colorectal Cancer
As opposed to the diversity metrics discussed above, OTU-based relative abundance profiles generated from 16S rRNA gene sequences are effective for classifying stool samples as originating from individuals with healthy, adenomatous, or cancerous colons [@Zackular:2014fba; @Baxter:2016dja]. By using classification models instead of attempting to identify single differentially abundant OTUs, these and other studies have been successful in capturing complex community relationships in which differences in taxonomic relative abundance are considered in the context of other taxa. The exceptional performance of bacteria in these classification models supports a role for bacterial functionality in colorectal cancer. We built off of these findings by evaluating the ability of virus community signatures to classify stool samples and compared their performance to models built using bacterial community signatures.

@@

The association between the bacterial and viral communities and colorectal cancer was driven by a few important microbes. *Fusobacterium* was the primary driver of the bacterial association with colorectal cancer, which is consistent with its previously described oncogenic potential **(Figure \ref{predmodel} C)**[@Flynn:2016iu]. The virome signature also was driven by a few OVUs, suggesting a role for these viruses in tumorigenesis **(Figure \ref{predmodel} D)**. It is also important to note that while these viruses were driving the signature, the magnitude of their importance and the significance of those values was noticibly less than the bacterial 16S signature, suggesting that unlike what is observed in the bacteria, there are many viruses that are associated with the cancerous state. The identified viruses were bacteriophages, belonging to *Siphoviridae*, *Myoviridae*, and phage taxa that could not be confidently identified beyond their broad phage identification (i.e. "unclassified"). Many of the important viruses, that were confirmed to not have genomic similarity to known bacterial genomes, were unidentifiable (denoted "unknown"). This is common in viromes across habitats; studies have reported as much as 95% of virus sequences belonging to unknown genomic units [@Pedulla:2003tu; @Hannigan:2015fz; @Willner:2009dq; @Brum:2015iaa]. When the bacterial and viral community signatures were combined, both bacterial and viral organisms drove the community association with cancer **(Figure \ref{predmodel} E)**.

## Phage Influence Between CRC Stages
Because previous work has identified shifts in which bacteria were most important at different stages of colorectal cancer [@Zackular:2014fba; @Zackular:2016en; @Flynn:2016iu], we explored whether shifts in the relative influence of phages could be detected between healthy, adenomatous, and cancerous colons. We evaluated community shifts between the disease stage transitions (healthy to adenomatous and adenomatous to cancerous) by building random forest models to compare only the diagnosis groups around the transitions. While bacterial OTU models performed equally well for all disease class comparisons, the virome model performances differed **(Figure \ref{transitionmodels} A-B)**. Like bacteria **(Figure \ref{transitionmodels} F-H)**, different virome members were important between the healthy to adenomatous and adenomatous to cancerous stages **(Figure \ref{transitionmodels} C-E)**.

@@

## Phage Dominance in CRC Virome
Differences in the colorectal cancer virome could have been driven by eukaryotic (human) viruses or by bacteriophages. To better understand the types of viruses that were important for colorectal cancer, we identified the virome OVUs as being similar to either eukaryotic viruses or bacteriophages. The most important viruses to the classification model were identified as bacteriophages (**Figure \ref{threewaymodel})**. Overall, we were able to identify `r signif(idcount, digits = sigfig)`% of the OVUs as known viruses, and `r signif(phagecount, digits = sigfig)`% of those viral OVUs aligned to bacteriophage reference genomes. It is important to note that this could have been influenced by our methodological biases against enveloped viruses (more common of eukaryotic viruses than bacteriophage), due to chloroform and DNase treatment for purification.

@@

## Community Context of Influential Phages
Because the link between colorectal cancer and the virome was driven by bacteriophages (as opposed to non-bacterial viruses), we tested a potential hypothesis that the virome signal was a mere reflection of the bacterial signal, and thus highly correlated with the bacterial signal. If this hypothesis were true, we would expect a correlation between the relative abundances of influential bacterial OTUs and virome OVUs. Instead, we observed a strikingly low correlation between bacterial and viral relative abundances **(Figure \ref{correlations} A,C)**. Overall, there was an absence of correlation between the most influential OVUs and bacterial OTUs **(Figure \ref{correlations} B)**. This evidence supported our null hypothesis that the influential viral OVUs were not primarily reflections of influential bacteria.

Given these findings, we posited that the most influential phages were acting by infecting a wide range of bacteria in the overall community, instead of just the influential bacteria. In other words, we hypothesized that the influential bacteriophages were community hubs (i.e. central members) within the bacteria and phage interactive network. We investigated the potential host ranges of all phage OVUs using a previously developed random forest model that relies on sequence features to predict which phages infected which bacteria in the community **(Figure \ref{network} A)** [@Hannigan:2018gv]. The predicted interactions were then used to identify phage community hubs. We calculated the alpha centrality (i.e. measure of importance in the ecological network) of each phage OVU's connection to the rest of the network. The phages with high centrality values were defined as community hubs. Next, the centrality of each OVU was compared to its importance in the colorectal cancer classification model. Phage OVU centrality was significantly and positively correlated with importance to the disease model (p-value `r pvalcalc(corstats[c(corstats$names %in% "pval"),"scores"])`, R = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`), suggesting that phages that were important in driving colorectal cancer also were more likely to be community hubs **(Figure \ref{network} B)**. Together these findings supported our hypothesis that influential phages were hubs within their microbial communities and had broad host ranges.

@@

Based on our findings, we have developed a conceptual model to be tested in our future studies aimed at elucidating the role the colonic virome plays in colorectal cancer **(Figure \ref{modelsummary} A)**. We found that basic diversity metrics of alpha diversity (richness and Shannon diversity) and beta diversity (Bray-Curtis dissimilarity) were insufficient for identifying virome community differences between healthy and cancerous states. By implementing a machine learning approach (random forest classification) to leverage inherent, complex patterns not detected by diversity measures, we were able to detect strong associations between the colon virus community composition and colorectal cancer. The dsDNA virome of colorectal cancer was composed primarily of bacteriophages. These phage communities were not exclusively predators of the most influential bacteria, as demonstrated by the lack of correlation between the abundances of the bacterial and phage populations. Instead, we identified influential phages as being community hubs, suggesting phages influence cancer by altering the greater bacterial community instead of directly modulating the influential bacteria. Our previous work has shown that modifying colon bacterial communities alters colorectal cancer progression and tumor burden in mice [@Zackular:2016en; @Baxter:2014hb]. This provides a precedent for phages indirectly influencing colorectal cancer progression by altering the bacterial community composition. Overall, our data support a model in which the bacteriophage community modulates the bacterial community, and through those interactions indirectly influences the bacteria driving colorectal cancer progression **(Figure \ref{modelsummary} A)**. Although our evidence suggested phages indirectly influenced colorectal cancer development, we were not able to rule out the role of phages directly interacting with the human host [@Lengeling:2013ia; @Grski:2012fa].

@@

In addition to the diagnostic ramifications for understanding the colorectal cancer microbiome, our findings suggest that viruses, while understudied and currently under-appreciated in the human microbiome, are likely to be an important contributor to human disease. Viral community dynamics have the potential to provide an abundance of information to supplement those of bacterial communities. Evidence has suggested that the virome is a crucial component to the microbiome and that bacteriophages are important players. Bacteriophage and bacterial communities cannot maintain stability and co-evolution without one another [@RodriguezValera:2009cr; @Brockhurst:2013iq]. Not only is the human virome an important element to consider in human health and disease [@Ly:2014ew; @Monaco:2016ita; @Willner:2009dq; @Abeles:2015dy; @Modi:2013fi; @SantiagoRodriguez:2015gd; @Norman:2015kb], but our findings support that it is likely to have a significant impact on cancer etiology and progression.

>>>>

Human viruses (those that infect human cells) have been associated with many cancers, largely due to their mutagenic and functionally manipulative abilities. Despite this, cancer microbiome studies have almost exclusively focused on bacteria instead of viruses. We began evaluating the cancer virome by focusing on colorectal cancer, a primary cause of morbidity and mortality throughout the world and a cancer linked to altered colonic bacterial community compositions but with an unknown association with the gut virome. We used 16S rRNA gene, whole shotgun metagenomic, and purified virus metagenomic sequencing of stool to evaluate the differences in human colorectal cancer virus and bacterial community composition. Through random forest modeling we identified differences in the healthy and colorectal cancer virome. The cancer-associated virome consisted primarily of temperate bacteriophages that were also predicted to be bacteria-virus community network hubs. These results provide foundational evidence that bacteriophage communities are associated with colorectal cancer and potentially impact cancer progression by altering the bacterial host communities.

@@

Colorectal cancer is a leading cause of cancer-related death in the United States and worldwide. Its risk and severity have been associated with colonic bacterial community composition. Although human-specific viruses have been linked to other cancers and diseases, little is known about colorectal cancer virus communities. We addressed this knowledge gap by identifying differences in colonic virus communities in the stool of colorectal cancer patients and how they compared to bacterial community differences. The results suggested an indirect role for the virome in colorectal cancer by modulating their associated bacterial community. These findings both support a biological role for viruses in colorectal cancer and provide a new understanding of basic colorectal cancer etiology.

@@

# Introduction
The human gut virome is the community of all viruses found in the gut, including bacteriophages (i.e. viruses that only infect bacteria), eukaryotic viruses (i.e. viruses that only infect eukaryotic cells), and human-specific viruses (i.e. viruses that only infect human cells). Due to their mutagenic abilities and propensity for functional manipulation, human viruses are strongly associated with, and in many cases cause, cancer [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up]. Because bacteriophages are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk], and members of those bacterial communities have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb; @Arthur:2012kl], bacteriophages have the potential to indirectly impact cancer as well. The gut virome therefore has a potential to be associated with, and potentially impact, human cancer. Altered human virome composition and diversity have already been identified in diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], cystic fibrosis [@Willner:2009dq], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb]. The strong association of bacterial communities with colorectal cancer, the previous identification of human-specific viruses that cause cancer, and the precedent for the virome to impact other human diseases suggest that colorectal cancer may be associated with altered virus communities.

Colorectal cancer is the second leading cause of cancer-related deaths in the United States [@Siegel:2014jo]. The US National Cancer Institute estimates that over 1.5 million Americans were diagnosed with colorectal cancer in 2016 and over 500,000 Americans died from the disease [@Siegel:2014jo]. Growing evidence suggests that an important component of colorectal cancer etiology may be perturbations in the colonic bacterial community [@Zackular:2016en; @Zackular:2014fba; @Baxter:2014hb; @Dejea:2014fz; @Arthur:2012kl]. Work in this area has led to a proposed disease model in which bacteria colonize the colon, develop biofilms, promote inflammation, and enter an oncogenic synergy with the cancerous human cells [@Flynn:2016iu]. This association also has allowed researchers to leverage bacterial community signatures as biomarkers to provide accurate, noninvasive colorectal cancer detection from stool [@Zackular:2014fba; @Baxter:2016dja; @Zeller:2014ix; @Sze2018; @Shah2017]. While an understanding of colorectal cancer bacterial communities has proven fruitful both for disease classification and for identifying the underlying disease etiology, bacteria are only a subset of the colon microbiome. Viruses are another important component of the colon microbial community that have yet to be studied in the context of colorectal cancer.

Colorectal cancer progresses in a stepwise process that begins when healthy tissue develops into a precancerous polyp (i.e., adenoma) in the large intestine [@Fearon:2011go]. If not removed, the adenoma may develop into a cancerous lesion that can invade and metastasize, leading to severe illness and death. Progression to cancer can be prevented when precancerous adenomas are detected and removed during routine screening [@Levin:2008kz; @Zauber:2015cr]. Survival for colorectal cancer patients may exceed 90% when the lesions are detected early and removed [@Levin:2008kz]. Thus, work that aims to facilitate early detection and prevention of progression beyond early cancer stages has great potential to inform therapeutic development. We evaluated disruptions in virus and bacterial community composition in a human cohort whose stool was sampled at the three relevant stages of cancer development: healthy, adenomatous, and cancerous.

Here we begin to address the knowledge gap of whether virus community composition is altered in colorectal cancer and, if it is, how those differences might impact cancer progression and severity. We also aimed to evaluate the virome's potential as a reservoir of diagnostic biomarkers. The implications of this study are threefold. *First*, this work supports a biological role for the virome in colorectal cancer development and suggests that more than the bacterial members of the associated microbial communities are involved in the process. *Second*, we present a supplementary virus-based approach for classification modeling of colorectal cancer using stool samples. *Third*, we provide initial support for the importance of studying the virome as a component of the microbiome ecological network, especially in cancer.

@@

## Unaltered Diversity in Colorectal Cancer
Microbiome and disease associations are often described as being of an altered diversity (i.e. "dysbiotic"). Therefore, we first evaluated the influence of colorectal cancer on virome OVU diversity. We evaluated differences in communities between disease states using the Shannon diversity, richness, and Bray-Curtis metrics. We observed no significant alterations in either Shannon diversity or richness in the diseased states as compared to the healthy state **(Figure \ref{betaogu} C-D)**. There was no statistically significant clustering of the disease groups (ANOSIM p-value `r pvalcalc(alphadf[c(alphadf$Title %in% "anovawithoutneg"),"Stat"])`, **Figure \ref{betaogu}**). Notably, there was a significant difference between the few blank controls that remained after rarefying the data and the other study groups (ANOSIM p-value `r pvalcalc(alphadf[c(alphadf$Title %in% "anovawithneg"),"Stat"])`, **Figure \ref{betaogunegative})**, further supporting the quality of the sample set. In summary, standard alpha and beta diversity metrics were insufficient for capturing virus community differences between disease states **(Figure \ref{betaogu})**. This is consistent with what has been observed when the same metrics were applied to 16S rRNA gene sequences and metagenomic samples [@Zeller:2014ix; @Zackular:2014fba; @Baxter:2016dja; @Sze2018; @Shah2017] and points to the need for alternate approaches to detect the impact of colorectal cancer disease state on these community structures.

## Virome Composition in Colorectal Cancer
As opposed to the diversity metrics discussed above, OTU-based relative abundance profiles generated from 16S rRNA gene sequences are effective for classifying stool samples as originating from individuals with healthy, adenomatous, or cancerous colons [@Zeller:2014ix; @Zackular:2014fba; @Baxter:2016dja; @Sze2018; @Shah2017]. By using classification models instead of attempting to identify single differentially abundant OTUs, these and other studies have been successful in capturing complex community relationships in which differences in taxonomic relative abundance are considered in the context of other taxa. The exceptional performance of bacteria in these classification models supports a role for bacterial functionality in colorectal cancer. We built off of these findings by evaluating the ability of virus community signatures to classify stool samples and compared their performance to models built using bacterial community signatures.

@@

The association between the bacterial and viral communities and colorectal cancer was driven by a few important microbes. *Fusobacterium* was the primary driver of the bacterial association with colorectal cancer, which is consistent with its previously described oncogenic potential **(Figure \ref{predmodel} C)**[@Flynn:2016iu]. The virome signature also was driven by a few OVUs, suggesting a role for these viruses in tumorigenesis **(Figure \ref{predmodel} D)**. It is also important to note that while these viruses were driving the signature, the magnitude of their importance and the significance of those values was noticeably less than the bacterial 16S rRNA gene signature, suggesting that unlike what is observed among the bacteria, there are many viruses that are associated with the cancerous state. These viruses were bacteriophages, belonging to *Siphoviridae*, *Myoviridae*, and phage taxa that could not be confidently identified beyond their broad phage identification (i.e. "unclassified"). Many of the important viruses, which were confirmed to not have genomic similarity to known bacterial genomes, were unidentifiable (denoted "unknown"). This is common in viromes across habitats; studies have reported as much as 95% of virus sequences belonging to unknown genomic units [@Pedulla:2003tu; @Hannigan:2015fz; @Willner:2009dq; @Brum:2015iaa]. When the bacterial and viral community signatures were combined, both bacterial and viral organisms drove the community association with cancer **(Figure \ref{predmodel} E)**.

## Phage Influence Between CRC Stages
Because previous work has identified shifts in which bacteria were most important at different stages of colorectal cancer, we explored whether shifts in the relative influence of phages could be detected between healthy, adenomatous, and cancerous colons. We evaluated community shifts between the disease stage transitions (healthy to adenomatous and adenomatous to cancerous) by building random forest models to compare only the diagnosis groups around the transitions. Although bacterial OTU models performed equally well for all disease class comparisons, the virome model performances varied across comparisons **(Figure \ref{transitionmodels} A-B)**. Like bacteria **(Figure \ref{transitionmodels} F-H)**, different virome members were important between the healthy to adenomatous and adenomatous to cancerous stages **(Figure \ref{transitionmodels} C-E)**.

@@

## Phage Dominance in CRC Virome
Differences in the colorectal cancer virome could have been driven by eukaryotic (human) viruses or by bacteriophages. To better understand the types of viruses that were important for colorectal cancer, we identified the virome OVUs as being similar to either eukaryotic viruses or bacteriophages. The viruses that were most important to the performance of the classification model were identified as bacteriophages (**Figure \ref{threewaymodel})**. Overall, we were able to identify `r signif(idcount, digits = sigfig)`% of the OVUs as known viruses, and `r signif(phagecount, digits = sigfig)`% of those viral OVUs aligned to bacteriophage reference genomes. It is important to note that this could have been influenced by our methodological biases against enveloped viruses (more common of eukaryotic viruses than bacteriophage), due to chloroform and DNase treatment for purification.

@@

## Community Context of Influential Phages
Because the link between colorectal cancer and the virome was driven by bacteriophages (as opposed to non-bacterial viruses), we tested the hypothesis that the virome signal was a mere reflection of the bacterial signal, and thus highly correlated with the bacterial signal. If this hypothesis were true, we would expect a correlation between the relative abundances of influential bacterial OTUs and virome OVUs. Instead, we observed a strikingly low correlation between bacterial and viral relative abundances **(Figure \ref{correlations} A,C)**. Overall, there was an absence of correlation between the most influential OVUs and bacterial OTUs **(Figure \ref{correlations} B)**. This evidence supported our null hypothesis that the influential viral OVUs were not primarily reflections of influential bacteria.

Given these findings, we hypothesize that the most influential phages act by infecting a wide range of bacteria in the overall community, instead of just the influential bacteria. In other words, we hypothesized that the influential bacteriophages were community hubs (i.e. central members) within the bacteria and phage interactive network. We investigated the potential host ranges of all phage OVUs using a previously developed random forest model that relies on sequence features to predict which phages infected which bacteria in the community **(Figure \ref{network} A)** [@Hannigan:2018gv]. The predicted interactions were then used to identify phage community hubs. We calculated the alpha centrality (i.e. measure of importance in the ecological network) of each phage OVU's connection to the rest of the network. The phages with high centrality values were defined as community hubs. Next, the centrality of each OVU was compared to its importance in the colorectal cancer classification model. Phage OVU centrality was significantly and positively correlated with importance to the disease model (p-value `r pvalcalc(corstats[c(corstats$names %in% "pval"),"scores"])`, R = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`), suggesting that phages that were important in driving colorectal cancer also were more likely to be community hubs **(Figure \ref{network} B)**. Together these findings supported our hypothesis that influential phages were hubs within their microbial communities and had broad host ranges.

@@

Based on our findings, we have developed a conceptual model to be tested in our future studies aimed at elucidating the role the colonic virome plays in colorectal cancer **(Figure \ref{modelsummary} A)**. We found that basic diversity metrics of alpha diversity (richness and Shannon diversity) and beta diversity (Bray-Curtis dissimilarity) were insufficient for identifying virome community differences between healthy and cancerous states. By implementing a machine learning approach to leverage inherent, complex patterns not detected by diversity measures, we were able to detect strong associations between the colon virus community composition and colorectal cancer. The dsDNA virome of colorectal cancer was composed primarily of bacteriophages. These phage communities were not exclusively predators of the most influential bacteria, as demonstrated by the lack of correlation between the abundances of the bacterial and phage populations. Instead, we identified influential phages as being community hubs, suggesting phages influence cancer by altering the greater bacterial community instead of directly modulating the influential bacteria. Our previous work has shown that modifying colon bacterial communities alters colorectal cancer progression and tumor burden in mice [@Zackular:2016en; @Baxter:2014hb]. This provides a precedent for phages indirectly influencing colorectal cancer progression by altering the bacterial community composition. Overall, our data support a model in which the bacteriophage community modulates the bacterial community, and through those interactions indirectly influences the bacteria driving colorectal cancer progression **(Figure \ref{modelsummary} A)**. Although our evidence suggested phages indirectly influenced colorectal cancer development, we were not able to rule out the role of phages directly interacting with the human host [@Lengeling:2013ia; @Grski:2012fa].

@@

In addition to the diagnostic ramifications for understanding the colorectal cancer microbiome, our findings suggest that viruses, while understudied and currently under-appreciated in the human microbiome, are likely to be an important contributor to human disease. Viral community dynamics have the potential to provide an abundance of information to supplement those of bacterial communities. The growing body of evidence suggests that the virome is a crucial component to the microbiome and that bacteriophages are important players. Bacteriophage and bacterial communities cannot maintain stability and co-evolution without one another [@RodriguezValera:2009cr; @Brockhurst:2013iq]. Not only is the human virome an important element to consider in human health and disease [@Ly:2014ew; @Monaco:2016ita; @Willner:2009dq; @Abeles:2015dy; @Modi:2013fi; @SantiagoRodriguez:2015gd; @Norman:2015kb], but our findings support that it is likely to have a significant impact on cancer etiology and progression.
